Constraints on Cosmic Rays, Magnetic Fields, and Dark Matter from Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS and \Fermi

### 2011-12-17 ms.tex

\begin{abstract}
\linenumbers
Observations of radio halos and relics in galaxy clusters indicate efficient electron acceleration
processes; the same processes are even more efficient for protons. Since protons have a cooling
time which exceeds the Hubble time in clusters, they can accumulate suggesting that clusters may
also be sources of very high-energy (VHE; $E>100$ GeV) gamma-ray emission resulting from the
interactions of highly relativistic cosmic-ray (CR) hadrons with the intracluster gas. Reported
here are VHE gamma-ray observations of the Coma galaxy cluster with the VERITAS array of imaging
Cherenkov telescopes, with complementing \Fermi Gamma-ray Space Telescope observations at GeV
energies. No significant gamma-ray emission from the Coma cluster was detected. Flux upper limits
at the 99\% confidence level were measured to be on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$
s$^{-1}$ (VERITAS, $>$220 GeV) and $\sim 3\times 10^{-6}$ ph. m$^{-2}$ s$^{-1}$ (\Fermi, 1-3
GeV), respectively. For VERITAS (\Fermi), the limits are a factor of 60 (20) higher than
model-predicted minimum gamma-ray fluxes, which assume the radio-emitting electrons of the Coma
radio halo to be all secondaries from cosmic-ray interactions. We use the gamma-ray flux upper
limits to constrain the populations of CRs and magnetic fields in Coma. Using an analytical
approach informed by cosmological cluster simulations, the CR-to-thermal pressure is constrained
to be $< 15\%$ from VERITAS data and $< 2.5\%$ from \Fermi data (averaged within the virial
radius). Assuming that the radio-emitting electrons of the Coma halo result from hadronic CR
interactions, the observations imply a lower limit on the central magnetic field in Coma of
$1$--$3\mu$G, depending on how the magnetic field declines with radius and on the gamma-ray
spectral index. Since these values are well below those inferred by Faraday rotation measurements
in Coma, this makes the hadronic model a very plausible explanation of the Coma radio
halo. Finally, a constraint on the thermally averaged product of the total self-annihilation cross
section and the relative velocity of the dark-matter particles, $\expval{\sigma v}$, is derived
and found to be $\lesssim 10^{-21}$ cm$^{3}$ s$^{-1}$ for $m_{\chi}\gtrsim 1000$ GeV/c$^{2}$.
\end{abstract}

>>>>

\begin{abstract}
\linenumbers
Observations of radio halos and relics in galaxy clusters indicate efficient electron acceleration
processes; the same processes are even more efficient for protons. Since protons have a cooling
time which exceeds the Hubble time in clusters, they can accumulate suggesting that clusters may
also be sources of very high-energy (VHE; $E>100$ GeV) gamma-ray emission resulting from the
interactions of highly relativistic cosmic-ray (CR) hadrons with the intracluster gas. Reported
here are VHE gamma-ray observations of the Coma galaxy cluster with the VERITAS array of imaging
Cherenkov telescopes, with complementing \Fermi Gamma-ray Space Telescope observations at GeV
energies. No significant gamma-ray emission from the Coma cluster was detected. Flux upper limits
at the 99\% confidence level were measured to be on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$
s$^{-1}$ (VERITAS, $>$220 GeV) and $\sim 3\times 10^{-6}$ ph. m$^{-2}$ s$^{-1}$ (\Fermi, 1-3
GeV), respectively. For VERITAS (\Fermi), the limits are a factor of 60 (20) higher than
model-predicted minimum gamma-ray fluxes, which assume the radio-emitting electrons of the Coma
radio halo to be all secondaries from cosmic-ray interactions. We use the gamma-ray flux upper
limits to constrain the populations of CRs and magnetic fields in Coma. Using an analytical
approach informed by cosmological cluster simulations, the CR-to-thermal pressure is constrained
to be $< 16\%$ from VERITAS data and $< 2.5\%$ from \Fermi data (averaged within the virial
radius). Assuming that the radio-emitting electrons of the Coma halo result from hadronic CR
interactions, the observations imply a lower limit on the central magnetic field in Coma of
$1$--$3\mu$G, depending on how the magnetic field declines with radius and on the gamma-ray
spectral index. Since these values are well below those inferred by Faraday rotation measurements
in Coma, this makes the hadronic model a very plausible explanation of the Coma radio
halo. Finally, a constraint on the thermally averaged product of the total self-annihilation cross
section and the relative velocity of the dark-matter particles, $\expval{\sigma v}$, is derived
and found to be $\lesssim 10^{-21}$ cm$^{3}$ s$^{-1}$ for $m_{\chi}\gtrsim 1000$ GeV/c$^{2}$.
\end{abstract}
